---
output:
  pdf_document: default
  html_document: default
---

# (PART\*) R {.unnumbered}


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide", message = FALSE, warning = FALSE, comment = "")
```

# Sandwich Estimator {#Sandwich-Estimator}

## Sandwich Estimator - R Function

<!-- The sandwich estimator unlike the modified Box correction does not use a REML estimate. Therefore we do not need to use a package for mixed models such as **lme4** or **nlme**. -->

The code below gives an R function that returns the estimate of $\beta$ using the GEE approach (denoted as $\hat\beta_W$ in paper I) under the setup of a multivariate linear model as described by equation \@ref(eq:general-model). The function also returns the sandwich estimator for the covariance matrix of the estimate of $\beta$. 

To use this function you need to copy the code (can click on clipboard symbol on top right corner of code chunk) and run it in R so that it is in your R environment. How to use this function is then illustrated by an example.

This `sandwich_estimator()` function takes four inputs: `model`, `subjects`, `data` and `W`. The `model` is used to describe the mean model and takes the same form as you would give a formula in `lm()`. The `subjects` variable is used to identify the variable in the data that labels the subjects. `W` is then used to set the working covariance matrix. In paper II the working covariance matrix is denoted as ($W^{-1}$).

```{r}
library(Matrix)

sandwich_estimator = function(model, subjects, data, W = NA){
  
  model_formula = as.formula(model)
  
  X = model.matrix(model_formula, data = data)
  full_model_frame = model.frame(model_formula, data = data)
  Y = model.response(full_model_frame)
  
if(is.na(W)){
  W = diag(1, nrow = nrow(X))
}
  
beta_W = solve(t(X)%*%solve(W)%*%X)%*%t(X)%*%solve(W)%*%Y
  
  
  sigma_i_comb = list()
  for(i in levels(data[[paste(subjects)]])){
    
    Y_i = Y[data[[subjects]] == i]
    X_i = X[data[[subjects]] == i,]
    sigma_i = (Y_i - (X_i %*% beta_W))
    sigma_i = sigma_i[,1]
    sigma_i = sigma_i %*% t(sigma_i)
    
    sigma_i_comb[[paste(i)]] = sigma_i
  }
  
  sigma = bdiag(sigma_i_comb)
  V_s = solve(t(X)%*%solve(W)%*%X)%*%t(X)%*%solve(W)%*%sigma%*%t(solve(t(X)%*%solve(W)%*%X)%*%t(X)%*%solve(W))
  
  return(list("beta_W" = beta_W, "V_s" = V_s, "sigma_i_comb" = sigma_i_comb))
  
}
```

## Wald Test using Sandwich Estimator - R Function

This `sandwich_estimator_p_value()` function carries out a Wald test using the sandwich estimate outputted from the `sandwich_estimator()` function, therefore both functions need to be in your R environment.

This function has an additional `model_r` variable, and you should assign to it a formula with the variables removed that you want to test for significance for in the full model defined in the `model` variable.

```{r}
sandwich_estimator_p_value = function(model, model_r, subjects, data, W = NA){
  
  sand_estimate = sandwich_estimator(model = model, subjects = subjects, data = data, W = W)

  Beta_r_names =
    colnames(model.matrix(model, data = data))[!(colnames(model.matrix(model, data = data)) %in%
  colnames(model.matrix(model_r, data = data)))]

  Beta_r = sand_estimate$beta_W[Beta_r_names,]
  
  Wald_sand_stat = (Beta_r %*% solve(sand_estimate$V_s[Beta_r_names,Beta_r_names]) %*% t(t(Beta_r)))[1]
  Wald_sand_stat_p_value = pchisq(q = Wald_sand_stat, df = length(Beta_r), lower.tail = FALSE)
  
return(list("Wald_sand_stat" = Wald_sand_stat, "Wald_sand_stat_p_value" = Wald_sand_stat_p_value, "Beta_r" = Beta_r))
  
}
```


***


### Example - Cardiac Enzyme Dataset {#Sandwich-Cardiac-Enzyme-Example}

In this example we look at the same dataset and mean model that was used in the example covered in Section 6.1 of paper II. We apply the `sandwich_estimator()` function to output the sandwich estimator of the covariance matrix of the estimated fixed effects. Then use the `sandwich_estimator_p_value()` to test for the inclusion of the interaction term between treatment and time in the model.


First of all we need to import the dataset. The code below imports the dataset from a GitHub repository, and converts a selection of the variables in the dataset to factors.

```{r}
Cardiac_enzyme_data = read.csv("https://raw.githubusercontent.com/DylanDijk/Simon-Skene-Mixed-Models-Tutorial/main/Data_Images_Figures/The%20Cardiac%20Enzyme%20Data%20-%20Reduced%20Data%20set.csv")
Cardiac_enzyme_data$dog = as.factor(Cardiac_enzyme_data$dog)
Cardiac_enzyme_data$trt = as.factor(Cardiac_enzyme_data$trt)
Cardiac_enzyme_data$time = as.factor(Cardiac_enzyme_data$time)
```

The code below runs the `sandwich_estimator` function:

```{r}
Cardiac_enzyme_sandwich_estimate = 
  sandwich_estimator(model = atp ~  trt + time + trt:time,
                     subjects = "dog", data = Cardiac_enzyme_data)

```

After running this function we can then extract the estimate of $\beta$ (`beta_W`) and the sandwich estimate of $\Sigma$ (`V_s`).

```{r}
Cardiac_enzyme_sandwich_estimate$beta_W
Cardiac_enzyme_sandwich_estimate$V_s
```

The code below runs the `sandwich_estimator_p_value()` function:

```{r}
Cardiac_enzyme_sandwich_wald_p_value = 
sandwich_estimator_p_value(model = atp ~ trt + time + trt:time,
                           model_r = atp ~ trt + time, 
                           data = Cardiac_enzyme_data, subjects = "dog")
```

And we get the following p-value:
```{r, results='markup'}
Cardiac_enzyme_sandwich_wald_p_value$Wald_sand_stat_p_value
```

***

## Adjusted Sandwich Estimator

The `adj_sandwich_estimator()` function below computes the Wald $F$-statistic as described in Section 3 of paper I. The function uses the sandwich estimator computed by the `sandwich_estimator()` function and therefore requires both of the previous functions to be in your R environment.

This function has the same input variables as the `adj_sandwich_estimator_p_value()` function.

```{r}
adj_sandwich_estimator = function(model, model_r, subjects, data, W = NA){
  
  sand_estimate = sandwich_estimator(model = model, subjects = subjects, data = data)
  
  sand_stat = sandwich_estimator_p_value(model = model, model_r = model_r, subjects = subjects, data = data)
  
  l = length(sand_stat$Beta_r)

K = list(length = l^2)

for(i in 1:l^2){
  
          vec = rep(0, l^2)

  if(((i - 1) %% (l + 1)) == 0){
    
      for(j in 1:l^2){
      
        if(((j - 1)) %% (l + 1) == 0){
          vec[j] = 1
          
        }
      }  
    }

     K[[i]] =  vec     
          
}

Beta_r_names = names(sand_stat$Beta_r)

K = do.call(rbind, K)
omega_hat = (diag(l^2) + K) %*% 
  (sand_estimate$V_s[Beta_r_names, Beta_r_names] %x%
    sand_estimate$V_s[Beta_r_names, Beta_r_names]
  )

X = model.matrix(model, data = data)

if(is.na(W)){
W = diag(1, nrow = nrow(X))
}

mod_based_est = solve(t(X) %*% solve(W) %*% X)

P_i = list()
 for(i in levels(data[[subjects]])){
    
    X_i = X[data[[subjects]] == i,]
    P_i[[i]] =
      c(
    t(X_i) %*% W[data[[subjects]] == i,data[[subjects]] == i] %*% sand_estimate$sigma_i_comb[[i]] %*%
    W[data[[subjects]] == i,data[[subjects]] == i] %*% X_i
      )
  }

n = nrow(data)

P_hat = Reduce("+", P_i)/n

P_i_minus_P_hat = lapply(P_i, function(x){x - P_hat})

T_hat = Reduce("+",lapply(P_i_minus_P_hat, function(x){x %*% t(x)}))/(n*(n-1))

psi_hat = n^2 * (mod_based_est[Beta_r_names,] %x% mod_based_est[Beta_r_names,]) %*%
             T_hat %*% t(mod_based_est[Beta_r_names,] %x% mod_based_est[Beta_r_names,])

v = sum(diag(psi_hat %*% omega_hat))/sum(diag(psi_hat %*% psi_hat))

F_Wald_adj_sand = (v - l + l)/(v * l) * sand_stat$Wald_sand_stat
  
F_Wald_adj_sand_p = pf(df1 = l, df2 = v - l +1, q = F_Wald_adj_sand, lower.tail = FALSE)

return(list("F_Wald_adj_sand" = F_Wald_adj_sand,
            "F_Wald_adj_sand_p" = F_Wald_adj_sand_p))

}
```

### Example

This example is analogue to the calculation of the p-value in subsection \@ref(Sandwich-Cardiac-Enzyme-Example). 

```{r, cache=TRUE}
Cardiac_enzyme_adj_sandwich_wald_p_value = 
adj_sandwich_estimator(model = atp ~ trt + time + trt:time,
                           model_r = atp ~ trt + time, 
                           data = Cardiac_enzyme_data, subjects = "dog")
```

We get the following Wald $F$-statistic and p-value:

```{r, results='markup'}
Cardiac_enzyme_adj_sandwich_wald_p_value$F_Wald_adj_sand
Cardiac_enzyme_adj_sandwich_wald_p_value$F_Wald_adj_sand_p
```


